Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-11-27 manuscript/manuscript.tex

We define an external computation point $P(r, \phi, \lambda)$ in a geocentric spherical
coordinate system at a radius $r$, geocentric latitude $\phi$, and longitude $\lambda$
where the gravitational fields are going to be calculated.
The first and second derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first and second derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary continuous function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.

>>>>

Consider an arbitrary tesseroid in a geocentric spherical coordinate system defined by
pairs of geocentric latitudinal ($\phi_1$, $\phi_2$), longitudinal ($\lambda_1$,
$\lambda_2$) and radial ($r_1$, $r_2$) boundaries.
We define an external computation point $P(r, \phi, \lambda)$ located at radius $r$,
geocentric latitude $\phi$, and longitude $\lambda$ where the gravitational fields the
tesseroid generates are going to be calculated.
The first- and second-order derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first- and second-order derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary continuous function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid
\citep{Uieda2016}.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.
